|

### 2021-08-10 prova_knit_finale.Rmd
little change
cambio APF

\subsubsection{Convergence of the IS estimator}\label{pf_is_conv}
We can show that the auto-normalized IS estimator is both consistent and asymptotically normal.\
Consider the auto-normalized IS estimator $\hat \varphi= \sum_{n=1}^N\varphi (X^n)W^n$. We can write:
\begin{equation*}
   \hat \varphi_{AN} - \E_{\Q}(\varphi)=\E_{\Q^N}(\varphi)-\E_{\Q}(\varphi)=\frac{N^{-1}\sum_{n=1}^Nw(X^n)[\varphi(X^n)-\E_{\Q}(\varphi)]}{N^{-1}\sum_{n=1}^Nw(X^n)}
\end{equation*}
Let $\bar \varphi(X):=\varphi(X)-\E_{\Q}(\varphi)$. Applying the CLT to the numerator, assuming that $\E_{\M}(w^2\bar\varphi^2)<+\infty$, we get:
\begin{equation}
    \sqrt{N}\Bigg\{\frac{1}{N}\sum_{n=1}^Nw(X^n)\bar \varphi (X^n)\Bigg\}\implies \N \big(0,\E_{\M}(w^2\bar\varphi^2)\big)\label{is_conv_clt}
\end{equation}
where $\implies$ denotes convergence in distribution. Instead, for the denominator the strong LLN implies that:
\begin{equation}
    \frac{1}{N}\sum_{n=1}^Nw(X^n)\xrightarrow{\text{a.s.}} \E_{\M}(w) \label{is_conv_lln}
\end{equation}
Applying Slutsky's theorem, together with \eqref{is_conv_clt} and \eqref{is_conv_lln}, we obtain:
\begin{equation}
   \sqrt{N}\big[  \hat \varphi_{AN} - \E_{\Q}(\varphi)\big] \implies \N \Bigg(0,\frac{\E_{\M}(w^2\bar\varphi^2)}{[\E_{\M}(w)]^2}\Bigg)
\end{equation}
Therefore, the IS (auto-normalized) estimator for $\varphi$ is asymptotically normal. Consistency can be proved in an analogous way, applying the strong LLN and lemma \ref{lemma_imp_sam_1}:
\begin{equation}
    \frac{1}{N}\sum_{n=1}^Nw(X^n) \varphi (X^n)\xrigtharrow{\text{a.s.}}  \E_\M(\varphi \cdot w)=\E_\Q(\varphi)\cdot \E_\M(w)
\end{equation}
Together with \eqref{is_conv_lln}, we get strong convergence of the estimator (that implies consistency).
\begin{equation}
   \hat \varphi_{AN}=\sum_{n=1}^NW^n \varphi (X^n)\xrightarrow{\text{a.s.}}  \E_\Q(\varphi)
\end{equation}

@@

The \textit{auxiliary particle filter} (APF) constitutes a further extension of the BPF. The GPF allowed to sample from a transition kernel different from the correct ones. The APF, instead, allows to extract the ancestor variables from an arbitrary distribution at the resampling step.\ \
Let $\eta_t:\X\rightarrow \R_+$ be a non-negative, real-valued positive function, called \textit{auxiliary function}. At time $t$, when the resampling stage occurs, the multinomial distribution for the indexes to be extracted can use probabilities that differ from the weights in $t-1$, $W_{t-1}^n$. The new weights, called \textit{auxiliary weights}, are computed as a transformation of the original ones, denoted as \textit{inferential weights}, through the auxiliary function:
\begin{equation}
    \tilde W_t^n=\frac{W_t^n\cdot \eta_t(X_t^n)}{\sum_{m=1}^NW_t^m\cdot \eta_t(X_t^n)}
\end{equation}
Note that the computation of the auxiliary weights is done from the auto-normalized inferential ones. Additionally, note that imposing $\eta_t(x_t)=1$ we recover the GPF and the BPF as special cases of the APF.\
Clearly, inferential weights are updated by incorporating the fact that resampling occurs according to different weights (the auxiliary ones):
\begin{equation}
    w_t^n=\frac{W_{t-1}^{A_t^n}}{\tilde W_{t-1}^{A_t^n}}G_t \label{apf_weights_up_}
\end{equation}
where $G_t$ is the incremental weight defined in the section on GPF.\ \

@@

It is possible to derive an optimality result for the auxiliary functions\footnote{Details in Chopin and Papaspiliopoulos (2020) th 10.2, p. 148}. In the special case in which the transition kernel used for sampling corresponds with the optimal one $M_t^*$, the optimal auxiliary function (i.e., the auxiliary function that minimizes the variance of the inferential weights) becomes:

@@

\item Normalize the auxiliary weights, for $n=1,...,N$:  $\tilde W_t^n:=\frac{\tilde w_t^n}{\sum_{m=1}^N\tilde w_t^m}$.\footnote{It is worth mentioning that the original APF included a final resampling step at the end of each iteration. However, our formulation is more efficient since having two resampling steps might introduce additional Monte Carlo variance. See Johansen and Evers (2007), p. 120}

@@

However, it is worth noting that the optimality results both for the kernel in the GPF and of the auxiliary  in the APF, has  two main weaknesses:

@@

    \item it is not clear that what is optimal at time $t$ remains such at future stages. In other words, it is not truly clear how the minimization of the variance of the inferential weights at each stage affects the efficiency of the PF estimators;

@@

For illustration purposes, we are going implement an auxiliary particle filter for the linear Gaussian model with auxiliary function $g(x_{t-1})=E(x_{t}|x_{t-1})=x_{t-1}$.

@@

We also implement the auxiliary particle filter with optimal transition kernel. Therfore the algorithm is similiar to the one discussed above with the difference that:

>>>>

\subsubsection{Convergence of the IS estimator}\label{pf_is_conv}
We can show that the auto-normalized IS estimator is both consistent and asymptotically normal.\
Consider the auto-normalized IS estimator $\hat \varphi= \sum_{n=1}^N\varphi (X^n)W^n$. We can write:
\begin{equation*}
   \hat \varphi_{AN} - \E_{\Q}(\varphi)=\E_{\Q^N}(\varphi)-\E_{\Q}(\varphi)=\frac{N^{-1}\sum_{n=1}^Nw(X^n)[\varphi(X^n)-\E_{\Q}(\varphi)]}{N^{-1}\sum_{n=1}^Nw(X^n)}
\end{equation*}
Let $\bar \varphi(X):=\varphi(X)-\E_{\Q}(\varphi)$. Applying the CLT to the numerator, assuming that $\E_{\M}(w^2\bar\varphi^2)<+\infty$, we get:
\begin{equation}
    \sqrt{N}\Bigg\{\frac{1}{N}\sum_{n=1}^Nw(X^n)\bar \varphi (X^n)\Bigg\}\implies \N \big(0,\E_{\M}(w^2\bar\varphi^2)\big)\label{is_conv_clt}
\end{equation}
where $\implies$ denotes convergence in distribution. Instead, for the denominator the strong LLN implies that:
\begin{equation}
    \frac{1}{N}\sum_{n=1}^Nw(X^n)\xrightarrow{\text{a.s.}} \E_{\M}(w) \label{is_conv_lln}
\end{equation}
Applying Slutsky's theorem, together with \eqref{is_conv_clt} and \eqref{is_conv_lln}, we obtain:
\begin{equation}
   \sqrt{N}\big[  \hat \varphi_{AN} - \E_{\Q}(\varphi)\big] \implies \N \Bigg(0,\frac{\E_{\M}(w^2\bar\varphi^2)}{[\E_{\M}(w)]^2}\Bigg)
\end{equation}
Therefore, the IS (auto-normalized) estimator for $\varphi$ is asymptotically normal. Consistency can be proved in an analogous way, applying the strong LLN and lemma \ref{lemma_imp_sam_1}:
\begin{equation}
    \frac{1}{N}\sum_{n=1}^Nw(X^n) \varphi (X^n)\xrightarrow{\text{a.s.}}  \E_\M(\varphi \cdot w)=\E_\Q(\varphi)\cdot \E_\M(w)
\end{equation}
Together with \eqref{is_conv_lln}, we get strong convergence of the estimator (that implies consistency).
\begin{equation}
   \hat \varphi_{AN}=\sum_{n=1}^NW^n \varphi (X^n)\xrightarrow{\text{a.s.}}  \E_\Q(\varphi)
\end{equation}

@@

The \textit{auxiliary particle filter} (APF) constitutes a further extension of the BPF. The GPF allowed to sample from a transition kernel different from the correct ones. Instead, the APF allows to extract the ancestor variables from an arbitrary distribution at the resampling step.\ \
Let $\eta_t:\X\rightarrow \R_+$ be a non-negative, real-valued positive function, called \textit{auxiliary function}. At time $t$, when the resampling stage occurs, the multinomial distribution for the indexes to be extracted can use probabilities that differ from the weights in $t-1$, $W_{t-1}^n$. The new weights, called \textit{auxiliary weights}, are computed as a transformation of the original ones, denoted as \textit{inferential weights}, through the auxiliary function:
\begin{equation}
    \tilde W_t^n=\frac{W_t^n\cdot \eta_t(X_t^n)}{\sum_{m=1}^NW_t^m\cdot \eta_t(X_t^n)}
\end{equation}
Note that the computation of the auxiliary weights is done from the auto-normalized inferential ones. Additionally, note that imposing $\eta_t(x_t)=1$ we recover the GPF and the BPF as special cases of the APF.\
Clearly, inferential weights are updated by incorporating the fact that resampling occurs according to different weights (the auxiliary ones):
\begin{equation}
    w_t^n=\frac{W_{t-1}^{A_t^n}}{\tilde W_{t-1}^{A_t^n}}G_t \label{apf_weights_up_}
\end{equation}
where $G_t$ is the incremental weight defined in the section on GPF.\ \

@@

It is possible to derive an optimality result for the auxiliary functions.\footnote{Details in Chopin and Papaspiliopoulos (2020) th 10.2, p. 148.} In the special case in which the transition kernel used for sampling corresponds with the optimal one $M_t^*$, the optimal auxiliary function (i.e., the auxiliary function that minimizes the variance of the inferential weights) becomes:

@@

\item Normalize the auxiliary weights, for $n=1,...,N$:  $\tilde W_t^n:=\frac{\tilde w_t^n}{\sum_{m=1}^N\tilde w_t^m}$.\footnote{It is worth mentioning that the original APF included a final resampling step at the end of each iteration. However, our formulation is more efficient since having two resampling steps might introduce additional Monte Carlo variance. See Johansen and Evers (2007), p. 120.\label{footnote_apf}}

@@

However, it is worth noting that the optimality results both for the kernel in the GPF and of the auxiliary function in the APF, have  two main weaknesses:

@@

    \item it is not clear that what is optimal at time $t$ remains such at future stages. In other words, it is not truly clear how the minimization of the variance of the inferential weights at each stage affects the overall efficiency of the PF estimators;

@@

For illustration purposes, we are going implement an auxiliary particle filter for the linear Gaussian model with auxiliary function $g(x_{t-1})=E(x_{t}|x_{t-1})=x_{t-1}$.  \

Firstly, states are drawn from the true transition kernel, while a second resampling step is done at the end of each iteration, according to the original formulation of the algorithm (see footnote \ref{footnote_apf}).

@@

We also implement the auxiliary particle filter with optimal transition kernel. Note that we are also using the perfectly adapted auxiliary function and that we resample only once, at the beginning of each iteration. Therefore the algorithm is similar to the one discussed above with the difference that:
